### 2016-09-29 rebuttal.tex

\begin{document}

\includepdf{coverletter.pdf}

\section{Comments from Editor}

\begin{questions}
\question I found the \url{http://live.sympy.org/} site very convenient for trying things
out in SymPy, but it is only mentioned in the supplement, which might be
overlooked by readers. I suggest moving that section into the main paper, as
it provides a great way to play along with the examples while reading the
paper.
\begin{solution}
\label{editorpoint1}
Section (previously) 9 from the supplement was edited and moved to the introduction:
\begin{quote}
  All the examples in this paper can be tested on
  \href{http://live.sympy.org}{SymPy Live}, an online Python shell that uses
  the Google App Engine to execute SymPy code. SymPy Live is also integrated
  in the SymPy documentation at
  \href{http://docs.sympy.org}{http://docs.sympy.org}.
\end{quote}
\end{solution}
\question The Basic Usage section omits what I think is an important point:
how does one distinguish between evaluating exp(1) in Python and exp(1) in
SymPy? In other words, how are symbolic constants specified?
\begin{solution}
  Both the Python standard library and SymPy have an exp callable, so it depends
  which symbols are imported in your global namespace: the \texttt{exp}
  function from the \texttt{math} module or the \texttt{exp} class from \texttt{sympy}.
  In the later case a special symbolic constant will be produced.

### 2016-09-29 rebuttal.tex

We have added a sentence to the section~\ref{P-sec:core}, that shows symbolic
constants as an example of leaf nodes:
\begin{quote}
Symbols or symbolic constants, like $e$ or $\pi$, are other examples of
leaf nodes.
\begin{verbatim}
>>> exp(1)
E
>>> exp(1).args
()
>>> x.args
()
\end{verbatim}
\end{quote}
\end{solution}
\question The first
thing I tried after seeing Table 2 (simplication functions) did not work:

\begin{verbatim}
>>> trigsimp (exp( Matrix(2, 2, [0, -y, y, 0])) )
\end{verbatim}
fails to recognize cos and
sin. Am I expecting too much?
\begin{solution}
  This indeed does not work. The implementation in \verb|trigsimp| primarily
  thus far has focused on transforming trigonometric functions into other
  trigonometric functions. The ability to simplify complex exponentials is
  something that we would like to work.
  \href{https://github.com/sympy/sympy/issues/11459}{Issue 11459} in our
  public issue tracker tracks this feature.
\end{solution}
\question Section 3.6: are eigenvalues and singular
values included? If not, why not?
\begin{solution}
Yes, they are included.  The second example in that section addresses the computation of
eigenvalues.  Singular values can be computed with
\texttt{A.singular\_values()}. We have not included this example, as the
singular value expressions for our example matrix \texttt{A} are quite large,
but we have added ``singular values'' to the list of features in the second
paragraph of the section.
\end{solution}
\question Like one of the referees, I expected to
see MPFR mentioned in Section 4.1, at least to mention the pros and cons and
say why it isn't used.
\begin{solution}

  The major advantage is that mpmath is a pure Python library. It's slower
  than MPFR, but not too much to evaluate any reasonably complicated single
  expression at interactive speed, which covers most use cases in SymPy. We
  have added a sentence about this to the numerics section.

  We've also mentioned that another advantage is that mpmath is BSD licensed,
  like SymPy, which is important for the scientific Python ecosystem (MPFR is
  licensed under the LGPL).
\end{solution}
\question Section 4.1: please state whether the syntax for
functions in mpmath is identical, or not, to that in Sympy.
\begin{solution}
The mpmath library is not a SymPy submodule, but a separate library,
which we have indicated in the text. Most function names in mpmath are the same as SymPy, but it is not true in every case. We have
added an example to the text showing an instance where they diverge
(summations).
\end{solution}
\question Unless I have
missed something, a weakness of the paper is that speed is mentioned only
twice, on lines 496 and 666. My assumption is that SymPy is slow compared with
its commercial competitors. Please comment on the speed issue, and not just in
the conclusions.
\begin{solution}
Section~\ref{P-sec:performance}, titled ``Performance'' was added to the manuscript, which discusses the
speed issue and lays out a path forward. The conclusion was simplified.
\end{solution}
\question Editors are needed for [20].
\begin{solution}
We have done this.
\end{solution}
\end{questions}

### 2016-09-29 rebuttal.tex

\section{Reviewer 1}
\subsection{Basic reporting}
\begin{questions}
\question Generally good. A part that confused me is the assertion (footnote 3) that ``If
$A$ and $B$ are Symbols created with \texttt{commutative=False} then SymPy will keep
$A\cdot B$
and $B\cdot A$ distinct.'' Does that mean that BOTH of them must be created this way,
and that $A$ and $x$ (if $x$ is created normally) will commute? Is there any way to
declare commutators? How does one guarantee that other pieces of code, e.g.
Gaussian elimination, respect non-commutativity?
\begin{solution}
  We have clarified the footnote. Both expressions must be set as
  \texttt{commutative=False}. Internally, in \texttt{Mul}, the ``commutative
  part'' of an expression is pulled out to the front and canonically ordered,
  and the ``noncommutative part'' is not reordered.

Commutators are implemented in the \texttt{sympy.physics.quantum} module and
are discussed in section~\ref{P-sec:quantum}.

Different algorithms do generally require special consideration for
noncommutative expressions to be correct. If an algorithm implicitly assumes
that expressions commute, it may return incorrect results when given a
noncommutative expression.
\end{solution}

\question References reasonable, though [8], an excellent reference, makes almost
precisely the opposite point to that for which it is cited---``simplification
is not well-defined''. I think the authors are trying to follow [8]'s
definition of simplification.
\begin{solution}
We agree that this is a bad reference, because we are not actually
using their definition.  Yes, that article rigorously defines
``simplification'', but it's just one approach.  This citation
was replaced with a better reference [34], that illustrates our point.
\end{solution}
\end{questions}

@@

\question Lines 151--153
Minor comment: srepr is a useful command. As someone who also uses Mathematica, I wonder if there is a sympy version of the TreeForm command which produces a visualisation of the expression tree, or maybe an output format that I could pass to a graph library for visualisation?

\begin{solution}
We have added a footnote about the \texttt{dotprint} function (footnote~\ref{P-note:dotprint}),
which outputs expressions in the dot format that can be rendered with Graphviz.
\end{solution}

\question Section 2.3: Assumptions
Is there a way of the user listing all available assumptions?

\begin{solution}
  To access this in code, the only way is
\begin{verbatim}
from sympy.core.assumptions import _assume_defined
\end{verbatim}
  but as this is a private variable (it begins with an underscore), it is not
  considered part of the public API, so we do not wish to recommend it in the
  paper. The recommended list is in the SymPy documentation, at
  \url{http://docs.sympy.org/latest/modules/core.html#module-sympy.core.assumptions}.
  We have opened \href{https://github.com/sympy/sympy/issues/11539}{issue 11539}
  to track this in our issue tracker.
\end{solution}
\question Section 2.4
Line 216: This is due in part because the same language, Python, is used both for the internal implementation and the external usage by users.

### 2016-09-29 rebuttal.tex

  \begin{enumerate}
  \item We have renamed the ``Features'' section to ``Overview of
    Capabilities''.
  \item We have renamed the ``Domain Specific Submodules'' section to
      ``Physics Submodule'' (see point~\ref{rev3point30}).
  \item We have moved the ``Architecture'' section to the end, after the
    ``Physics Submodule'' section.
  \item We have moved the ``Basic Usage'' subsection to the beginning of the
    ``Overview of Capabilities'' section, and added an intro paragraph to the
    ``Overview of Capabilities'' section.
  \item We have moved the ``Assumptions'' subsection to the ``Overview of
    Capabilities'' subsection, after the features table.
  \item The ``Comparison with Mathematica'' section has been moved to the end of
    the supplement.
  \item The ``Other Projects that
      Depend on SymPy'' section (which has been renamed from ``Other Projects that
      Use SymPy'') has been moved from the supplement to the main paper content.
  \end{enumerate}

We disagree with the reviewer on a few points. We do not consider SymPy Live
or SymPy Gamma to be key components of SymPy. The goal of the paper is
to be about the SymPy library. While SymPy Live and SymPy Gamma are
maintained by the SymPy community, they are separate projects. A
mention of SymPy Live has been added to the introduction (see editor
point~\ref{editorpoint1}), as it may assist an interested reader in trying the
examples from the paper. A mention of SymPy Gamma has been added to the ``Projects that
Depends on SymPy'' table (Table~\ref{P-projects-table}) (see below). We have
left the more in-depth discussion of SymPy Gamma in the supplement.

@@

\item The ``Statistics'' module is not a core module and is independent (no
  other SymPy submodules depend on it).
\end{itemize}

The ``Architecture'' section is a key component of the paper. The goal of the
paper is to discuss the architecture and core features of SymPy. It therefore
would be inappropriate to move this section to the supplement. We have
followed the reviewer's advice and moved the architecture section to the end
of the paper, to make the paper easier to follow.
\end{solution}
\end{questions}

\begin{questions}

\question Explaining what Python is (lines 70--72) should go before talking about SymPy as a ``CAS written in Python''. Further, the paper assumes a moderate familiarity with Python (for example, Python's console, OOP, and exceptions), and this should be specified. There should be a short note on the used Python console (\verb|>>>| is the prompt, with the results of computation following immediately in the lines after it). The citation [25] from line 65 should be moved next to ``Python'' in line 70.
\begin{solution}
The citation was moved and a footnote~\ref{P-note:prompt} for the Python interactive prompt was
  added where it was used first time. We have also added a footnote indicating
  a moderate familiarity with Python is assumed.
\end{solution}

### 2016-09-29 supplement.tex
Fix supplement ~ in front of \ref

The supplementary material take a deeper look at certain topics in SymPy for
which there was not enough room to discuss in the paper.
Section \ref{suppsec:Gruntz} discusses the Gruntz algorithm, used to
calculate limits in the SymPy.  Sections \ref{suppsec:Series}--\ref{suppsec:Ten}
discuss in depth some selected submodules.  Section \ref{suppsec:numsimpl}
discusses numerical simplification.  Section \ref{suppsec:examples} provides
additional examples for topics, discussed in the main paper.  In section
\ref{sympy-gamma} the SymPy Gamma
and the SymPy Live projects are introduced.  Finally, section \ref{comp-mma}
has a
brief comparison of SymPy with Wolfram Mathematica.

>>>>

The supplementary material take a deeper look at certain topics in SymPy for
which there was not enough room to discuss in the paper.
Section~\ref{suppsec:Gruntz} discusses the Gruntz algorithm, used to
calculate limits in the SymPy.  Sections~\ref{suppsec:Series}--\ref{suppsec:Ten}
discuss in depth some selected submodules.  Section~\ref{suppsec:numsimpl}
discusses numerical simplification.  Section~\ref{suppsec:examples} provides
additional examples for topics, discussed in the main paper.  In section
\ref{sympy-gamma} the SymPy Gamma
and the SymPy Live projects are introduced.  Finally, section \ref{comp-mma}
has a
brief comparison of SymPy with Wolfram Mathematica.
